Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-03-15 iso-int/rc-iso-int.tex
citations and collecting things for intro

Reversible languages

@@

Previous work on {{langRevT}} \cite{rc2011,infeffects} studied its
properties and various constructions in it. However, for a
computational model, there has been little focus on actually writing
programs directly in {{langRevT}}. This paper aims to remedy that by
presenting a systematic technique of expressing logically reversible
small-step abstract machines in {{langRevT}}. This paper shows that
once we have a logically reversible machine in a notation of our
choice, expressing the machine as an isomorphic interpreter in
{{langRevT}} can be done systematically and does not present any
significant conceptual difficulties.

@@

We briefly review the reversible language {{langRevT}} introduced in our
previous work~\cite{rc2011,infeffects}. The set of types includes the empty
type {{0}}, the unit type {{1}}, sum types {{b1+b2}}, product types
{{b1*b2}}, and recursive types {{@@x.b}}. The set of values {{v}} includes
{{()}} which is the only value of type {{1}}, {{left v}} and {{right v}}
which inject {{v}} into a sum type, {{(v1,v2)}} which builds a value of
product type, and {{<v>}} which is used to construct recursive values. There
are no values of type {{0}}:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b | x | @@x.b
% values, v ::= () | left v | right v | (v, v) | <v>

>>>>

Landscape:

- write program; derive inverse
program~\cite{Gluck:2005,Mu:2002:IFF:648086.757330,conf/aplas/GluckK03,Gluck:2005};
we're not interested in this; we want to directly write partial isomorphisms

 - design reversible abstract machines by hand (e.g.,
 SE(M)CD~\cite{Kluge:1999:SEMCD} or SECD-H~\cite{lorenz}) and then implement
 them; this is what we are doing; our implementation is in this language in
 which every computation is a partial isomorphism so the implementation will
 be checked to be reversible by construction

 - generally one can take any reversible algorithm; transform it to a
 small-step reversible abstract machine, and then implement the machine in
 our
 language~\cite{Danvy:2008:RRN:1813347.1813350,Rendel:2010:ISD:1863523.1863525}.

 \cite{Yokoyama:2008:PRP,Axelsen:2011:RPC:1987171.1987176} argue that the
 clean simulation of reversible Turing machines is a criterion for computing
 strength of reversible languages. This is kind of what we show.

 \cite{Yokoyama:2007:RPL:1244381.1244404} also show a reversible
 self-interpreter which is what we show

 we argue and so does \cite{Rendel:2010:ISD:1863523.1863525} that partial iso
 are a good foundation for reversible programming

@@

Previous work on {{langRevT}} \cite{rc2011,James:2012:IE:2103656.2103667}
studied its properties and various constructions in it. However, for a
computational model, there has been little focus on actually writing programs
directly in {{langRevT}}. This paper aims to remedy that by presenting a
systematic technique of expressing logically reversible small-step abstract
machines in {{langRevT}}. This paper shows that once we have a logically
reversible machine in a notation of our choice, expressing the machine as an
isomorphic interpreter in {{langRevT}} can be done systematically and does
not present any significant conceptual difficulties.

@@

We briefly review the reversible language {{langRevT}} introduced in our
previous work~\cite{rc2011,James:2012:IE:2103656.2103667}. The set of types
includes the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
product types {{b1*b2}}, and recursive types {{@@x.b}}. The set of values
{{v}} includes {{()}} which is the only value of type {{1}}, {{left v}} and
{{right v}} which inject {{v}} into a sum type, {{(v1,v2)}} which builds a
value of product type, and {{<v>}} which is used to construct recursive
values. There are no values of type {{0}}:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b | x | @@x.b
% values, v ::= () | left v | right v | (v, v) | <v>
